---
title: "Drought NMDS"
author: "Kelly Endres"
date: "January 23, 2019"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls())
setwd("/Users/kelly/OneDrive/Desktop/Kelly/Abundance/Ipo")
```


```{r setup, include=FALSE}
#load packages
library(vegan)
library(dplyr)
library(tidyverse)
library(reshape2)
library(GGally)
```


```{r setup, include=FALSE}

#inport data
ipo1997 <- read.csv("data_1997.csv") #1997
ipo1998 <- read.csv("data_1998.csv") #1998
ipo2001 <- read.csv("data_2001.csv") #2001
ipo2012 <- read.csv("data_2012.csv") #2012
ipo2018 <- read.csv("data_2018.csv") #2018
```

To complete NMDS, the data from all years must be together on the same matrix. This requires loading the correct data and formatting it the same. In particular, the visitor column must be in the same order. 


```{r}
##1997
#treat year as factor
ipo1997$Year <- as.factor(ipo1997$Year)
#create dataframe with site, visitor, and total visits
ipo1997 <- cbind.data.frame(mdtsite = ipo1997$Site, mdtvisitor = ipo1997$Visitor, total.vis = ipo1997$Total.Visits)
#organize data by site then visitor
data_tvis_1997 <- dcast(ipo1997, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(data_tvis_1997)[1] <- "1" #rename
data_tvis_1997 <- data_tvis_1997[, -1] #delete
#rename columns
row.names(data_tvis_1997) <- c("P", "T", "V")
#add missing columns
data_tvis_1997 <- cbind(data_tvis_1997, "Other fly" = c(0, 0, 0), "Nectar robber" = c(0, 0, 0), "Small insect" = c(0, 0, 0), "Wasp" = c(0, 0, 0))
#order columns
data_tvis_1997 <- data_tvis_1997[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Nectar robber", "Other fly", "Small insect", "Solitary bee", "Wasp")]
#add year
data_tvis_1997 <- cbind(data_tvis_1997, "Year" = c(1997, 1997, 1997))
#add drought
data_tvis_1997 <- cbind(data_tvis_1997, "Drought" = c("Non-drought", "Non-drought", "Non-drought"))

##1998
#treat year as factor
ipo1998$Year <- as.factor(ipo1998$Year)
#create dataframe with site, visitor, and total visits
ipo1998 <- cbind.data.frame(mdtsite = ipo1998$Site, mdtvisitor = ipo1998$Visitor, total.vis = ipo1998$Total.Visits)
#organize data by site then visitor
data_tvis_1998 <- dcast(ipo1998, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(data_tvis_1998)[1] <- "1" #rename
data_tvis_1998 <- data_tvis_1998[, -1] #delete
#rename columns
row.names(data_tvis_1998) <- c("P", "T", "V")
#add missing columns
data_tvis_1998 <- cbind(data_tvis_1998, "Butterfly" = c(0, 0, 0), "Nectar robber" = c(0, 0, 0), "Small insect" = c(0, 0, 0), "Wasp" = c(0, 0, 0))
#order columns
data_tvis_1998 <- data_tvis_1998[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Nectar robber", "Other fly", "Small insect", "Solitary bee", "Wasp")]
#add year
data_tvis_1998 <- cbind(data_tvis_1998, "Year" = c(1998, 1998, 1998))
#add drought
data_tvis_1998 <- cbind(data_tvis_1998, "Drought" = c("Non-drought", "Non-drought", "Non-drought"))

##2001
#treat year as factor
ipo2001$Year <- as.factor(ipo2001$Year)
#create dataframe with site, visitor, and total visits
ipo2001 <- cbind.data.frame(mdtsite = ipo2001$Site, mdtvisitor = ipo2001$Visitor, total.vis = ipo2001$Total.Visits)
#organize data by site then visitor
data_tvis_2001 <- dcast(ipo2001, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(data_tvis_2001)[1] <- "1" #rename
data_tvis_2001 <- data_tvis_2001[, -1] #delete
#rename columns
row.names(data_tvis_2001) <- c("P", "T", "V")
#removes Gnat  andda
data_tvis_2001 <- data_tvis_2001[ ,-c(2, 3)] 
#add missing columns
data_tvis_2001 <- cbind(data_tvis_2001, "Nectar robber" = c(0, 0, 0), "Small insect" = c(0, 0, 0), "Butterfly" = c(0, 0, 0))
#order columns
data_tvis_2001 <- data_tvis_2001[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Nectar robber", "Other fly", "Small insect", "Solitary bee", "Wasp")]
#add year
data_tvis_2001 <- cbind(data_tvis_2001, "Year" = c(2001, 2001, 2001))
#add drought
data_tvis_2001 <- cbind(data_tvis_2001, "Drought" = c("Non-drought", "Non-drought", "Non-drought"))

##2012
#treat year as factor
ipo2012$Year <- as.factor(ipo2012$Year)
#treat NAs as 1
ipo2012[is.na(ipo2012)] <- 1
#create dataframe with site, visitor, and total visits
ipo2012 <- cbind.data.frame(mdtsite = ipo2012$Site, mdtvisitor = ipo2012$Visitor, total.vis = ipo2012$Total.Visits)
#organize data by site then visitor
data_tvis_2012 <- dcast(ipo2012, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(data_tvis_2012)[1] <- "1" #rename
data_tvis_2012 <- data_tvis_2012[, -1] #delete
#rename columns
row.names(data_tvis_2012) <- c("Avery Picnic Slope", "Avery Picnic Upper", "Back of Baldy Bend", "Baldy Lunch", "Bellview Bench", "Bellview View", "billy's Hilly Lower", "billy's Hilly Upper", "Emerald Lake Lower", "Emerald Lake Upper", "End of Gothic Hill", "Kebler Pass", "Long Lake Parking Lot 1", "Long Lake Parking Lot 2", "Long Lake Parking Lot 3", "Long Lake Parking Lot 4", "Max Field Meadow Lower", "Max Field Meadow Upper", "Ol' Dusty Road", "Rosy Point Lower", "Rosy Point Upper", "Schofield/Vera Falls", "Slate River 2", "Up on dat hill", "Vera Falls", "Vera Falls 2")
#removes Gnat column
data_tvis_2012 <- data_tvis_2012[,-c(3)]
#add missing columns
data_tvis_2012 <- cbind(data_tvis_2012, "Nectar robber" = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0))
#order columns
data_tvis_2012 <- data_tvis_2012[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Nectar robber", "Other fly", "Small insect", "Solitary bee", "Wasp")]
#add year
data_tvis_2012 <- cbind(data_tvis_2012, "Year" = c(2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012))
#add drought
data_tvis_2012 <- cbind(data_tvis_2012, "Drought" = c("Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought"))

##2018
#treat year as factor
ipo2018$Year <- as.factor(ipo2018$Year)
#create dataframe with site, visitor, and total visits
ipo2018 <- cbind.data.frame(mdtsite = ipo2018$Site, mdtvisitor = ipo2018$Visitor, total.vis = ipo2018$Total.Visits)
#organize data by site then visitor
data_tvis_2018 <- dcast(ipo2018, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(data_tvis_2018)[1] <- "1" #rename
data_tvis_2018 <- data_tvis_2018[, -1] #delete
#rename columns
row.names(data_tvis_2018) <- c("Avery Slope", "Behind Rustler's Gultch", "Bellview Bench Park", "Bellview View", "East River Overlook", "Rosy Point", "Vera Falls")
#add missing columns
data_tvis_2018 <- cbind(data_tvis_2018, "Nectar robber" = c(0, 0, 0, 0, 0, 0, 0), "Solitary bee" = c(0, 0, 0, 0, 0, 0, 0))
#order columns
data_tvis_2018 <- data_tvis_2018[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Nectar robber", "Other fly", "Small insect", "Solitary bee", "Wasp")]
#add year
data_tvis_2018 <- cbind(data_tvis_2018, "Year" = c(2018, 2018, 2018, 2018, 2018, 2018, 2018))
#add drought
data_tvis_2018 <- cbind(data_tvis_2018, "Drought" = c("Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought"))
```


The next step is to load all years together onto the same matrix. This matrix only holds visitor data by year, not by site. 

```{r}
#combine data
all.years <- rbind(data_tvis_1997, data_tvis_1998, data_tvis_2001, data_tvis_2012, data_tvis_2018)

```

Next I need to remove the drought column from the matrix with all the other data and format it as the grouping value. This means I'll put it into a value and the remove it from all.years

```{r}
#create value with drought information
Drought <- c(all.years$Drought)

#remove drought column from all.years matrix
all.years2 <- all.years[, -c(10, 11)]

#reformat as matrix
all.years2 <- as.matrix(all.years2)
```

```{r}
# run a NMDS ordination
nmds <- metaMDS(all.years2)

plot(nmds, type = "t") 
  ordiellipse(nmds, Drought, col=1:4)

ordiplot(nmds,type="n")
orditorp(nmds,display="species",col="black",air=0.01, cex = .5)
orditorp(nmds,display="sites", col=c(rep("darkgreen", 9), rep("firebrick", 33)),cex=.5,air=0.01)
ordiellipse(nmds, Drought, draw = "polygon", col=c("darkseagreen1", "darksalmon")) 

```

Summary:

The NMDS plot shown in this figure visualizes the difference in space between drought and non-drought years. The two areas show some overlap intidating partial similarities in composition. Some sites fall away from the expexted drought and non-drought years, indiciating within year differences. 