---
title: "Abundance data"
author: "Kelly Endres"
date: "January 19, 2019"
output: html_document
---

This worksheet will be to look at relative species abundances. 

```{r}
rm(list = ls())
setwd("/Users/kelly/OneDrive/Desktop/Kelly/Abundance")

library(ggplot2)
```

First I will make a dataframe for each year with the data we need. This means creating a dataframe with year, visitor, and total visit information for each of the 9 grouped catagories - Bumblebee, Butterfly, Hummingbird, Hoverfly, Wasp, Small insect, Other fly, Nectar robber, Solitary bee - with 0 visits added for the groups that have no recorde visits in a year. 



```{r}

#1997

#import data
ipovar97 <- read.csv("data_1997.csv")

# remove all columns but Year, Visitor, and Total.Visits
ipovar97 <- ipovar97[, -c(2, 3, 4, 5, 6, 8, 10, 11, 12, 13, 14, 15, 16)]

#create new dataframe (a18) that has visits per visitor group
a97 <- aggregate(Total.Visits ~ ., ipovar97, sum)

#add line for the missing groups so they are correctly identified as 0
x97 <- as.data.frame(rbind(c(1997, "Nectar robber", 0))) #make dataframe
colnames(x97) <- c("Year", "Visitor", "Total.Visits") #rename columns to match
x97$Visitor <- as.factor(x97$Visitor) #make visitor column factor
a97<- rbind(a97, x97) #add line to first total dataframe
x971 <- as.data.frame(rbind(c(1997, "Wasp", 0))) #make dataframe
colnames(x971) <- c("Year", "Visitor", "Total.Visits") #rename columns to match
x971$Visitor <- as.factor(x971$Visitor) #make visitor column factor
a97<- rbind(a97, x971) #add line to first total dataframe
x972 <- as.data.frame(rbind(c(1997, "Other fly", 0))) #make dataframe
colnames(x972) <- c("Year", "Visitor", "Total.Visits") #rename columns to match
x972$Visitor <- as.factor(x972$Visitor) #make visitor column factor
a97 <- rbind(a97, x972) #add line to first total dataframe
x973 <- as.data.frame(rbind(c(1997, "Small insect", 0))) #make dataframe
colnames(x973) <- c("Year", "Visitor", "Total.Visits") #rename columns to match
x973$Visitor <- as.factor(x973$Visitor) #make visitor column factor
a97 <- rbind(a97, x973) #add line to first total dataframe
a97$Total.Visits <- as.numeric(a97$Total.Visits) #make Total.Visits column numeric
a97$Year <- as.integer(a97$Year) #make Year column integer

#create new column with relative species abundances
rsa97 <- a97$Total.Visits/sum(a97$Total.Visits) #calculate relative species abundance 
a97 <- cbind(a97, rsa97) #add column to main dataframe
```

```{r}
#1998

#import data
ipovar98 <- read.csv("data_1998.csv")

# remove all columns but Year, Visitor, and Total.Visits
ipovar98 <- ipovar98[, -c(2, 3, 4, 5, 6, 8, 10)]

#create new dataframe (a98) that has visits per visitor group
a98 <- aggregate(Total.Visits ~ ., ipovar98, sum)

#add line for the missing groups so they are correctly identified as 0
x98 <- as.data.frame(rbind(c(1998, "Nectar robber", 0))) #make dataframe
colnames(x98) <- c("Year", "Visitor", "Total.Visits") #rename columns to match
x98$Visitor <- as.factor(x98$Visitor) #make visitor column factor
a98<- rbind(a98, x98) #add line to first total dataframe
x981 <- as.data.frame(rbind(c(1998, "Wasp", 0))) #make dataframe
colnames(x981) <- c("Year", "Visitor", "Total.Visits") #rename columns to match
x981$Visitor <- as.factor(x981$Visitor) #make visitor column factor
a98<- rbind(a98, x981) #add line to first total dataframe
x982 <- as.data.frame(rbind(c(1998, "Other fly", 0))) #make dataframe
colnames(x982) <- c("Year", "Visitor", "Total.Visits") #rename columns to match
x982$Visitor <- as.factor(x982$Visitor) #make visitor column factor
a98 <- rbind(a98, x982) #add line to first total dataframe
x983 <- as.data.frame(rbind(c(1998, "Small insect", 0))) #make dataframe
colnames(x983) <- c("Year", "Visitor", "Total.Visits") #rename columns to match
x983$Visitor <- as.factor(x983$Visitor) #make visitor column factor
a98 <- rbind(a98, x983) #add line to first total dataframe
a98$Total.Visits <- as.numeric(a98$Total.Visits) #make Total.Visits column numeric
a98$Year <- as.integer(a98$Year) #make Year column integer

#create new column with relative species abundances
rsa98 <- a98$Total.Visits/sum(a98$Total.Visits) #calculate relative species abundance 
a98 <- cbind(a98, rsa98) #add column to main dataframe
```

```{r}
#2001

#import data
ipovar01 <- read.csv("data_2001.csv")

# remove all columns but Year, Visitor, and Total.Visits
ipovar01 <- ipovar01[, -c(2, 3, 4, 5, 6, 8, 10, 11)]

#create new dataframe (a98) that has visits per visitor group
a01 <- aggregate(Total.Visits ~ ., ipovar01, sum)
a01 <- a01[- c(2, 3), ] #remove Gnat? column and Hawkmoth --> TBD FIGURE OUT WHAT TO DO

#add line for the missing groups so they are correctly identified as 0
x01 <- as.data.frame(rbind(c(2001, "Butterfly", 0))) #make dataframe
colnames(x01) <- c("Year", "Visitor", "Total.Visits") #rename columns to match
x01$Visitor <- as.factor(x01$Visitor) #make visitor column factor
a01<- rbind(a01, x01) #add line to first total dataframe
x011 <- as.data.frame(rbind(c(2001, "Small insect", 0))) #make dataframe
colnames(x011) <- c("Year", "Visitor", "Total.Visits") #rename columns to match
x011$Visitor <- as.factor(x011$Visitor) #make visitor column factor
a01<- rbind(a01, x011) #add line to first total dataframe
x012 <- as.data.frame(rbind(c(2001, "Nectar robber", 0))) #make dataframe
colnames(x012) <- c("Year", "Visitor", "Total.Visits") #rename columns to match
x012$Visitor <- as.factor(x012$Visitor) #make visitor column factor
a01<- rbind(a01, x012) #add line to first total dataframe
row.names(a01) <- c(1, 2, 3, 4, 5, 6, 7, 8, 9) #rename rows
a01$Total.Visits <- as.numeric(a01$Total.Visits) #make Total.Visits column numeric
a01$Year <- as.integer(a01$Year) #make Year column integer

#create new column with relative species abundances
rsa01 <- a01$Total.Visits/sum(a01$Total.Visits) #calculate relative species abundance 
a01 <- cbind(a01, rsa01) #add column to main dataframe
```


```{r}
#2012

#import data
ipovar12 <- read.csv("data_2012.csv")

# remove all columns but Year, Visitor, and Total.Visits
ipovar12 <- ipovar12[, -c(2, 3, 4, 5, 6, 8)]
ipovar12[is.na(ipovar12)] <- 1 #treat nas in total visits as 1 (since there was only 1 visit the number of visits wasn't put down)

#create new dataframe (a12) that has visits per visitor group
a12 <- aggregate(Total.Visits ~ ., ipovar12, sum)
a12 <- a12[- c(3), ] #remove Gnat? column

#add line for the missing groups so they are correctly identified as 0
x12 <- as.data.frame(rbind(c(2012, "Nectar robber", 0))) #make dataframe
colnames(x12) <- c("Year", "Visitor", "Total.Visits") #rename columns to match
x12$Visitor <- as.factor(x12$Visitor) #make visitor column factor
a12<- rbind(a12, x12) #add line to first total dataframe
row.names(a12) <- c(1, 2, 3, 4, 5, 6, 7, 8,  9) #rename rows
a12$Total.Visits <- as.numeric(a12$Total.Visits) #make Total.Visits column numeric
a12$Year <- as.integer(a12$Year) #make Year column integer

#create new column with relative species abundances
rsa12 <- a12$Total.Visits/sum(a12$Total.Visits) #calculate relative species abundance 
a12 <- cbind(a12, rsa12) #add column to main dataframe

```


```{r}
#2018

#import data
ipovar18 <- read.csv("data_2018.csv")

# remove all columns but Year, Visitor, and Total.Visits
ipovar18 <- ipovar18[, -c(2, 3, 4, 5, 6)]

#create new dataframe (a18) that has visits per visitor group
a18 <- aggregate(Total.Visits ~ ., ipovar18, sum)

#add line for the missing groups so they are correctly identified as 0
x18 <- as.data.frame(rbind(c(2018, "Nectar robber", 0))) #make dataframe
colnames(x18) <- c("Year", "Visitor", "Total.Visits") #rename columns to match
x18$Visitor <- as.factor(x18$Visitor) #make visitor column factor
a18<- rbind(a18, x18) #add line to first total dataframe
x181 <- as.data.frame(rbind(c(2018, "Solitary bee", 0))) #make dataframe
colnames(x181) <- c("Year", "Visitor", "Total.Visits") #rename columns to match
x181$Visitor <- as.factor(x181$Visitor) #make visitor column factor
a18<- rbind(a18, x181) #add line to first total dataframe
a18$Total.Visits <- as.numeric(a18$Total.Visits) #make Total.Visits column numeric
a18$Year <- as.integer(a18$Year) #make Year column integer

#create new column with relative species abundances
rsa18 <- a18$Total.Visits/sum(a18$Total.Visits) #calculate relative species abundance 
a18 <- cbind(a18, rsa18) #add column to main dataframe
```

Now that each individual dataframe has been created they need to be merged together into one. 

```{r}
#rename last columns to rsa, dropping the year at the end
colnames(a97) <- c("Year", "Visitor", "Total.visits", "rsa")
names(a98) <- names(a97) 
names(a01) <- names(a97)
names(a12) <- names(a97)
names(a18) <- names(a97)

# merge dataframes
all <- rbind(a97, a98, a01, a12, a18)
# make year factor for easier graphing 
all$Year <- as.factor(all$Year)
```

Graphing

I'm making a simple bar plot for visualization. This is not analysis, which will be done with adonis (?) but helps to simplify the data. 

This graph shows the relative species abundances for each year. Not useful except for my own visualization. 

```{r}
#plot realtive species abundances
allplot <- ggplot(na.omit(all), aes(x=Year, y=rsa, fill=Visitor)) + #graph
    geom_bar(stat="identity") + 
    xlab("Year") + #x labels 
    ylab("Relative Species Abundance") + #y labels
    ggtitle("Abundance") #title

#clean up graph
done <- allplot + scale_fill_brewer(palette="BrBG", direction=-1) + #color
  theme(axis.text = element_text(size = 13)) + #axis text
  theme(axis.title = element_text(size = 15)) + #title text
  theme(plot.title = element_text(size = 20, face = "bold")) #bold

#print
print(done)
```

This graph shows the overall abundance of each species per year. 

```{r}
#plot species abundances
bigplot <- ggplot(na.omit(all), aes(x=Year, y=Total.visits, fill=Visitor)) + #graph
    geom_bar(stat="identity") + 
    xlab("Year") + #xaxis
    ylab("Number of Visits") + #yaxis
    ggtitle("Abundance") #title

#clean up graph
bigdone <- bigplot + scale_fill_brewer(palette="BrBG", direction=-1) + #color
  theme(axis.text = element_text(size = 13)) + #axis text
  theme(axis.title = element_text(size = 15)) + #title text
  theme(plot.title = element_text(size = 20, face = "bold")) #bold

#pint
print(bigdone)
```

